A portable, annotation-based, visual stepper for Common Lisp

### 2020-02-24 paper/paper.tex
Slightly more readable paper

From an end user's perspective, to put the new SLY/Emacs stepper to
work means pressing the key chord \texttt{C-c C-s P} (\emph{control-c,
  control-s, capital P}) while the cursor is on a top-level form.
This causes the interesting sub-forms of that top-level form to be
automatically decorated with sticker \emph{overlays}, which by default
uses different shades of the color gray.  As described in
\ref{stickers}, a posterior compilation of that same top-level form
shall \emph{arm} the stickers and convert the overlay's color into
shades of blue.  From this point the stickers are executed as soon as
the user arranges for the instrumented code to be run as usual.

Note that the default behavior of the underlying sticker system
doesn't equate the execution of an instrumented form to an invocation
of the Lisp debugger.  This is by design.  As was explained in section
\ref{stickers}, in the default configuration, executions of the
instrumented form merely record the its return values for later
replay, which can be achieved with the key chord \texttt{C-c C-s C-r}.
Alternatively, the key chord \texttt{C-c C-s S} can be used to fetch
the most recent recordings for each sticker and visually decorate the
source code, indicating (1) stickers that have been executed; (2)
those that haven't yet, and (3) those that have exited non-locally.

Finally to enable the classic stepper functionality, the user must
explicitly select ``breaking stickers'' by affecting the value of the
\texttt{SLYNK-STICKERS:*BREAK-ON-STICKERS*} variable.

@@

In section \ref{collecting-interesting-forms} we sidestepped a
notorious difficulty with atomic forms, i.e. one-symbol symbolic
expressions.  In fact, this class of difficulties is already alluded
to in \cite[I-4.8]{annotation-based}.  The problem can be observed
with the simplest of forms:

@@

In this example, we note that atom \texttt{X} has two different
\emph{manifestations} in the encompassing form.  Naturally one wants
only the latter to be annotated, and not the first.  However, this is
hard to determine reliably since both are represented by the very same
object.  This is in stark contrast to compound forms represented by
different \texttt{CONS} cells.

We can enhance the form/position pairings table used above to record
the fact that there is more than one manifestation of an atom, but
that's not enough to know in \texttt{MNESIC-MACROEXPAND-ALL} which of
those is a in a position of execution.  This is because the
\texttt{AGNOSTIC-LIZARD} macroexpander will only traverse sub-forms of
forms actually returned by a macro's expansion.  In this example, our
hook is only called on the \texttt{(lambda (x) x)} and \texttt{x} forms,
\emph{not} on the \texttt{(x)} form, thus there is no easy way to
invalidate the first manifestation.

However, since we \emph{do} know that \texttt{x} is manifested at
\texttt{(9 . 10)} and \texttt{(12 . 13)} it is possible to devise
heuristics to trace back to the knowledge gathered when first reading
the form and traverse the atom's parent forms, given only the atom.  A
simple heuristic can proceed like this: if the atom exists inside a
compound form that does not occur in the final expansion, then that
atom isn't interesting, otherwise, it is.  This appears to solve the
above situation but fails miserably in the presence of the
\texttt{LOOP} macro since this macro has all the variable definition
``unprotected'' by parenthesis.  Hard-wiring exceptions to
\texttt{LOOP} and other macros could ameliorate the situation, but
overall this strategy feels murky and insufficient.

@@

For now, the proposed SLY/Emacs's stepper works around this limitation
by behaving conservatively and only annotating atoms is positions that
are guaranteedly safe, such as in function calls, which makes for the
majority of situations in practice.  Furthermore, users can always
manually add stickers to other atom manifestations they are interested
in and know to also be safe.

On the other hand, if more ambitious strategies of annotation are
attempted, the SLY/Emacs sticker system has already shown to be
reliable in the sense that if it needs to fail, it will do so early,
and thus won't mislead users to wrong debugging conclusions.

@@

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual practical use.  This was the
thinking behind the SLY/Emacs stepper's annotation-based approach.  We
think it especially fortuitous that the manner of source-code
correlation in SLY/Emacs's stickers feature doesn't make use of direct
references to source files or file positions, As was shown in
\ref{stickers}, it merely keeps an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs.  This
mechanism is simple but effective, and adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or adding
white-space and comments.  This detail is crucial in making stickers
and stepping actually usable for day-to-day programming, since the
user isn't dragged away from his editor.  These very advantages of
annotation-based steppers were already hinted at back in 1989 by the
proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

@@

It may be also be noted that the usual stepping paradigm where the
program is stopped at each point is only one of the possibilities
afforded by stickers.  As we explained in sections \ref{stickers} and
\ref{results}, two further ones can constitute innovative means of
debugging: the \emph{post-mortem} replay of sticker recordings and the
visual decoration of source code with colors to indicate the most
recent state of each sticker's execution.

Its current limitations notwithstanding, the SLY/Emacs stepper
described in this essay is the nearest thing a portable, modern,
visual stepping facility available in the Common Lisp ecosystem.  In
the process we have also described (1) a relatively simple (and
near-portable) technique for constructing a source-tracking reader in
terms of a compliant reader implementation and (2) a reliable
technique to carry over source-tracking information to macroexpansion.
It is conceivable that other debugging tools could be constructed from
either of these elements.

>>>>

We have released the result of our work on the GitHub
platform\footnote{See
  \texttt{git@github.com:joaotavora/sly-slepper.git}.}.

From an end user's perspective, to put the new SLY/Emacs stepper to
work means pressing the key chord \texttt{C-c C-s P} (\emph{control-c,
  control-s, capital P}) while the cursor is on a top-level form.
This causes the interesting sub-forms of that top-level form to be
automatically decorated with sticker \emph{overlays}, which by default
uses different shades of the color gray.  As described in
\ref{stickers}, a posterior compilation of that same top-level form
shall \emph{arm} the stickers and convert the overlays' color to
shades of blue.  From this point on, the stickers are executed as soon
as the user arranges for the instrumented code to be run as usual.

Note that the default behavior of the sticker system doesn't equate
the execution of an instrumented form to a break point, i.e. the
invocation of the Lisp debugger.  This is by design.  As was explained
in section \ref{stickers}, the default behavior is to have sticker
executions merely record the return values (or non-local exits) for
later replay.  This which can be achieved with the key chord
\texttt{C-c C-s C-r} or via \texttt{M-x sly-stickers-replay}.
Alternatively, the key chord \texttt{C-c C-s S} (or \texttt{M-x
  sly-stickers-fetch}) can be used to fetch the most recent recordings
for each sticker and visually decorate the source code, indicating (1)
stickers that have been executed; (2) those that haven't yet, and (3)
those that have exited non-locally.

Finally, to enable the classic stepper functionality, the user must
explicitly select ``breaking stickers'' by affecting the value of the
\texttt{SLYNK-STICKERS:*BREAK-ON-STICKERS*} variable.

@@

In section \ref{collecting-interesting-forms}, we sidestepped a
notorious difficulty with atomic forms, i.e. one-symbol symbolic
expressions.  This class of difficulties is already alluded to in
\cite[I-4.8]{annotation-based}.  The problem with atoms can be
observed with the simplest of forms:

@@

In this example, we note that the atom \texttt{X} has two different
\emph{manifestations} in the encompassing form.  Naturally one wants
only the latter to be annotated, and not the first.  However, this is
hard to determine reliably since both are represented by the very same
object.  This is in stark contrast to compound forms represented by
different \texttt{CONS} cells.

We can enhance the form/position pairings table used above to record
the fact that there is more than one manifestation of an atom, but
that's not enough to know in \texttt{MNESIC-MACROEXPAND-ALL} which of
those is a in a position of execution.  This is because the
\texttt{AGNOSTIC-LIZARD} macroexpander will only traverse sub-forms of
forms actually returned by a macro's expansion.  In this example, our
hook is only called on the \texttt{(lambda (x) x)} and \texttt{x}
forms, \emph{not} on the \texttt{(x)} form.  The latter form is merely
an argument to the macro itself where we have no power of
intervention, and thus there is no easy way to invalidate the first
manifestation.

However, since we \emph{do} know that \texttt{x} is manifested at
\texttt{(9 . 10)} and \texttt{(12 . 13)} it is possible to devise
heuristics to trace back to the knowledge gathered when first reading
the form and traverse the atom's parent forms, given only the atom.  A
very simple heuristic can proceed like this: if the atom exists inside
a compound form that does not occur in the final expansion, then that
atom isn't interesting, otherwise, it is.  This appears to solve the
above situation but fails miserably in the presence of the
\texttt{LOOP} macro since this macro has all the variable definition
``unprotected'' by parenthesis.

Hard-wiring exceptions to \texttt{LOOP} and other macros could
ameliorate the situation, but overall this strategy feels murky and
insufficient.  On the other hand, if more aggressive strategies of
atomic annotation are attempted, the SLY/Emacs sticker system has
already shown to be reliable in the sense that if it needs to fail
(because of an incorrect form being annotated), it will mostly do so
early.  Thus the potential to mislead users to wrong debugging
conclusions is minimized.

@@

For now, the proposed SLY/Emacs's stepper works around this limitation
by behaving conservatively and only annotating atoms in positions that
are guaranteed to be safe, such as inside function call forms.  This
makes for the majority of situations in practice.  Furthermore, users
can always manually add stickers to other atom manifestations they are
interested in and know to also be safe.

@@

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual, practical use.  This was the
thinking behind the SLY/Emacs stepper.  We think it especially
fortuitous that annotation-based approach and the manner of
source-code correlation in SLY/Emacs's stickers feature doesn't make
use of direct references to source files or file positions.  As was
shown in \ref{stickers}, we merely keep an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs, a
mechanism that is simple but effective.  It adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or even
adding white-space and comments inside it.  This detail is crucial in
making stickers and stepping actually usable for day-to-day
programming, since the user isn't dragged away from his editor or
forced to a special confinement while stepping.  We note that these
advantages of annotation-based steppers were already hinted at back in
1989 by the proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

@@

It may also be noted that the usual stepping paradigm where the
program is stopped at each point is only one of the possibilities
afforded by stickers.  As we explained in sections \ref{stickers} and
\ref{results}, two further ones can constitute innovative means of
debugging: the \emph{post-mortem} replay of sticker recordings and the
visual decoration of source code with colors to indicate the most
recent state of each sticker's execution.

Its current limitations notwithstanding, the SLY/Emacs stepper
described in this essay is the nearest thing an effective, portable
and visual stepping facility available in the Common Lisp ecosystem.
In the process we have also described (1) a relatively simple (and
near-portable) technique for constructing a source-tracking reader in
terms of a compliant reader implementation and (2) a reliable
technique to carry over source-tracking information to macroexpansion.
It is conceivable that other debugging tools could be constructed from
either of these elements.
